Improved Hydrodynamic/Reaction Coupling for Flows With Nuclear Statistical Equilibrium

### 2020-10-29 simplified_sdc_intro/paper.tex
comments

We solve the Euler equations for compressible, reacting flow.  For ease
of exposition we describe the one-dimensional case;
multidimensional extensions are a straightforward modification to
include the CTU hydrodynamics scheme.
The system appears as:
\begin{equation}
\frac{\partial \Uc_t}{\partial t}  + \frac{\partial \Fb^\xv (\Uc)}{\partial x} = \Sc(\Uc)
\end{equation}
where
\begin{equation}
\Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho \alpha_l \\
           \rho u \\
           \rho E \end{array}\right )
\end{equation}
are the conserved fluid quantities: mass density, $\rho$, velocity,
$u$, specific total energy, $E$, and nuclear species mass fractions,
$X_k$, and auxiliary composition variables, $a_l$.  The mass fractions
are constrained to sum to 1, $\sum_k X_k = 1$, but no such constraint
exists on the auxiliary variables.  The specific total energy relates
to the specific internal energy as $E = e + u^2/2$.  The corresponding
fluxes are
\begin{equation}
\Fb^\xv (\Uc) = \left ( \begin{array}{c}
   \rho u \\
   \rho X_k u \\
   \rho \alpha_l u \\
   \rho u u + p \\
   \rho u E + u p
   \end{array} \right )
\end{equation}
Here the pressure, $p$, enters, and is found via the equation of state.  For a system
where the composition is completely specified by the mass fractions, $X_k$, the equation
of state would take the form:
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}
However, as we'll see shortly, when we use the NSE table, we are using
the results of a much larger network then can be represented by the
mass fractions, and in this case, we rely on the auxiliary composition
variables, $\alpha_l$, to describe the state of the composition, and
our EOS has the form:
\begin{equation}
p = p(\rho, \alpha_l, e)
\end{equation}

@@

The compositional quantities
it carries, $\bar{A}$ and $Y_e$ are not representable from the 19
isotopes we carry in the main network. For this reason, when we are
using the NSE network, we always take the composition quantities in
the EOS directly from the auxiliary state in each zone instead of
using the $X_k$ directly. Our equation of state needs $\bar{Z}$ in addition
to the auxiliary quantities, which is is easily computed as
\MarginPar{might be worth spelling out exactly what inputs the NSE EOS needs to make things crystal clear}
\begin{equation}
\bar{Z} = \bar{A} Y_e
\end{equation}

@@

For Strang splitting, the basic flow of using the aprox19 + NSE network
is as follows:
\begin{itemize}

@@

  The hydrodynamics update proceeds as normal, but with an advection
  equation for each of the auxiliary composition variables:
  \begin{align}
    \ddt{(\rho Y_e)} + \ddx{(\rho Y_e u)} &= 0 \\
    \ddt{(\rho \bar{A})} + \ddx{(\rho \bar{A} u)} &= 0 \\
    \ddt{[\rho (B/A)]} + \ddx{[\rho (B/A) u]} &= 0 
  \end{align}

@@

@@

        This is the approach used in \citet{ma:2013}.  We call the table with the 
        input state:
        \begin{equation}
          \nse{\rho, T, (Y_e)_{\rm in}} \rightarrow \bar{A}_\out, (X_k)_\out, (B/A)_\out, dY_e/dt
        \end{equation}

@@

        This starts out the same as the above method:
        \begin{equation}
          \nse{\rho, T, (Y_e)_{\rm in}} \rightarrow \bar{A}^\star, (X_k)^\star, (B/A)^\star, dY_e/dt
        \end{equation}
        but we indicate with a $\star$ that most of those values are
        provisional, and we will seek a better value in the corrector step.
        We only update $Y_e$ from this call:
        \begin{equation}
          (Y_e)_{\rm out} = (Y_e)_{\rm in} + \Delta t \frac{dY_e}{dt}
        \end{equation}
        and then we use the table again, but with the updated $Y_e$:
        \begin{equation}
          \nse{\rho, T, (Y_e)_\out} \rightarrow \bar{A}_\out, (X_k)_\out, (B/A)_\out, dY_e/dt
        \end{equation}
        This corrects the composition so that it is consistent with the updated
        $Y_e$.  We use these values $\bar{A}_\out$, $(X_k)_\out$, and $(B/A)_\out$ to then complete the update, computing the energy as in Eq.~\ref{eq:nse_energy}.  \MarginPar{we could imagine looping over the EOS to get an updated T and then the table to get the updated $\bar{A}$, to see if that converges?}

>>>>

We solve the Euler equations for compressible, reacting flow.  For ease
of exposition we describe the one-dimensional case;
multidimensional extensions are a straightforward modification to
include the CTU hydrodynamics scheme.
The system appears as:
\begin{equation}
\frac{\partial \Uc_t}{\partial t}  + \frac{\partial \Fb^\xv (\Uc)}{\partial x} = \Sc(\Uc)
\end{equation}
where
\begin{equation}
\Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho \alpha_l \\
           \rho u \\
           \rho E \end{array}\right )
\end{equation}
are the conserved fluid quantities: mass density, $\rho$, velocity,
$u$, specific total energy, $E$, and nuclear species mass fractions,
$X_k$, and auxiliary composition variables, $\alpha_l$.  The mass fractions
are constrained to sum to 1, $\sum_k X_k = 1$, but no such constraint
exists on the auxiliary variables.  The specific total energy relates
to the specific internal energy as $E = e + u^2/2$.  The corresponding
fluxes are
\begin{equation}
\Fb^\xv (\Uc) = \left ( \begin{array}{c}
   \rho u \\
   \rho X_k u \\
   \rho \alpha_l u \\
   \rho u u + p \\
   \rho u E + u p
   \end{array} \right )
\end{equation}
Here the pressure, $p$, enters, and is found via the equation of state.  For a system
where the composition is completely specified by the mass fractions, $X_k$, the equation
of state would take the form:
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}
However, as we'll see shortly, when we use the NSE table, we are using
the results of a much larger network then can be represented by the
mass fractions, and in this case, we rely on the auxiliary composition
variables, $\alpha_l$, to describe the state of the composition, and
our EOS has the form:
\begin{equation}
p = p(\rho, \alpha_l, e)
\end{equation}

@@

The compositional quantities
it carries, $\bar{A}$ and $Y_e$ are not representable from the 19
isotopes we carry in the main network. For this reason, when we are
using the NSE network, we always provide these two composition quantities as
inputs to the EOS rather than using the $X_k$ directly.
the EOS directly from the auxiliary state in each zone instead of
using the $X_k$ directly.
Our equation of state needs $\bar{Z}$ in addition
to the auxiliary quantities, which is is easily computed as
\begin{equation}
\bar{Z} = \bar{A} Y_e
\end{equation}

@@

For Strang splitting, the hydrodynamic and reactive substeps over
the overall time-advancemenet scheme using the aprox19 + NSE network
are as follows:
\begin{itemize}

@@

  At the beginning of each hydrodynamic update we have an input state,
  $\Uc_{\rm in}$ that we wish to integrate over $\Delta t$ to obtain $\Uc_{\rm out}$
  The hydrodynamics update proceeds as normal, but with an advection
  equation for each of the auxiliary composition variables:
  \begin{align}
    \ddt{(\rho Y_e)} + \ddx{(\rho Y_e u)} &= 0 \\
    \ddt{(\rho \bar{A})} + \ddx{(\rho \bar{A} u)} &= 0 \\
    \ddt{[\rho (B/A)]} + \ddx{[\rho (B/A) u]} &= 0 
  \end{align}

@@

  At the beginning of each reactive update we have an input state,
  $\Uc_{\rm in}$ that we wish to integrate over $\Delta t/2$ to obtain $\Uc_{\rm out}$.

@@

        This is the approach used in \citet{ma:2013}.  We call the table with the 
        input state:
        \begin{equation}
          \nse{\rho_{\rm in}, T_{\rm in}, (Y_e)_{\rm in}} \rightarrow \bar{A}_\out, (X_k)_\out, (B/A)_\out, dY_e/dt
        \end{equation}

@@

        This starts out the same as the above method:
        \begin{equation}
          \nse{\rho, T, (Y_e)_{\rm in}} \rightarrow \bar{A}^\star, (X_k)^\star, (B/A)^\star, dY_e/dt
        \end{equation}
        but we indicate with a $\star$ that most of those values are
        provisional, and we will seek a better value in the corrector step.
        We only update $Y_e$ from this call:
        \begin{equation}
          (Y_e)_{\rm out} = (Y_e)_{\rm in} + \Delta t \frac{dY_e}{dt}
        \end{equation}
        and then we use the table again, but with the updated $Y_e$
        (note that $\rho_{\rm out}=\rho_{\rm in}$ in the Strang reaction formulation):
        \MarginPar{Indicate what $T$ is used.}
        \begin{equation}
          \nse{\rho_{\rm out}, T_{\rm ???}, (Y_e)_\out} \rightarrow \bar{A}_\out, (X_k)_\out, (B/A)_\out, dY_e/dt
        \end{equation}
        This corrects the composition so that it is consistent with the updated
        $Y_e$.  We use these values $\bar{A}_\out$, $(X_k)_\out$, and $(B/A)_\out$ to then complete the update, computing the energy as in Eq.~\ref{eq:nse_energy}.  \MarginPar{we could imagine looping over the EOS to get an updated T and then the table to get the updated $\bar{A}$, to see if that converges?}
